---
title: "SF Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SF_Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>"
)
```


## Setup  
* Install the package and any necessary prereqs by running `devtools::install_github("cmhoove14/Rsynthpops")`  
* If you don't have one already, get a US census api key at https://api.census.gov/data/key_signup.html. To store your api key, run `tidycensus::census_api_key(your_key, install = TRUE)` replacing `your_key` with the key you obtained from the url above.   
* Choose your target geography. If you choose a state or region containing multiple states, have that/those state's two character state abbreviation(s) handy. If it's a county, have the county's 5-digit Federal Information Processing Standard (FIPS) code handy. These are easy to find via a web search with the county's name and "fips code".  

```{r setup}
devtools::load_all()
#library(Rsynthpops)
```

```{r target_geo}
state_use  = "CA"
fips_use   = "06075"
year_use   = 2019
survey_use = "acs5"
pumas_use  = cts_to_pumas %>% 
  # If fips_use is not a 5-digit county code, update filter for state (STATEFP) or census tract (CTFP)
  filter(STCNTYFP == fips_use) %>% 
  pull(PUMA5CE) %>% 
  unique()
```

## Main commands  
* `rsp_get_` commands to download raw PUMS, ACS, and census data  
* `rsp_process_` commands to convert raw data into household/group quarters and person level seed and target data  
* `rsp_synth_` commands to synthesize your population from your target and seed datasets  

## Key steps  

### Getting input data  
The first decision to be made is which household and person characteristics are to be included in the synthetic population. `Rsynthpops` builds on functions from `tidycensus` to get PUMS and ACS data and adds an additional function to get data on group quarters directly from the census ftp site. Make sure to get and/or supply your census api key in order to use the `tidycensus` download functions. Currently supported household and person level characteristics are maintained in the `pums_to_acs` list that ships with `Rsynthpops`.  

```{r get_data, cache=TRUE}
# Variables we want to include in the final synthetic population
characteristics <- c(
  "Sex",
  "Age",
  "Race",
  "Ethnicity",
  "Occupation",
  "Grade",
  "School_Type",
  "HH_Income",
  "HH_Type",
  "HH_Size"
)

# Get group quarters data
gq_dat <- rsp_get_gq(
  STATE = state_use,
  LEVEL = "Tract"
)

# Get PUMS data
pums_dat <- rsp_get_pums(
  VARS   = characteristics,
  SURVEY = "acs5",
  STATES = state_use,
  YEAR   = year_use
)

# Get ACS data
acs_dat <- rsp_get_acs(
  VARS   = characteristics,
  SURVEY = "acs5",
  STATES = state_use,
  LEVEL  = "Tract",
  YEAR   = year_use
)
```

### Processing data  
Now that all the data is downloaded, it needs to be processed to feed into the synthesizer, in this case, the `ipu()` function from the `ipfr` package. Other synthesizers are being considered and tested, but iterative proportional updating is currently the canonical method. `ipu()` requires "seed" and "target" data for every agent characteristics, where the seeds come from individual pums records and targets come from aggregated acs/census records.  

```{r process_setup}
all_pums_hh_vars <- pums_variables %>% 
  filter(year == year_use,
         survey == survey_use,
         level == "housing") %>% 
  pull(var_code) %>% 
  unique()

# Determine which variables are household level
pums_vars <- colnames(pums_dat)
hh_vars <- pums_vars[which(pums_vars %in% all_pums_hh_vars)]

# Determine which variables are person level as those leftover from household vars
p_vars <- pums_vars[which(!pums_vars %in% all_pums_hh_vars)]

# Determine fips of areas to synthesize population for
fips_pops <- acs_dat[[1]] %>% 
  filter(variable == "B01001_001",
         grepl(fips_use, GEOID)) %>% 
  group_by(GEOID) %>% 
  summarise(pop = sum(estimate)) %>% 
  filter(pop > 0)

synth_fips <- fips_pops %>% pull(GEOID)
```

#### Generate synthetic group quarters population  
This is a key step for most target geographies. Group quarters residents make up around 3% of the entire US population and make up a substantial proportion of the population in some census tracts, e.g. those that contain college campuses, military bases, prisons, etc. Populations can be synthesized excluding the group quarters population, but will likely overestimate the number of households in the target geography and have higher error rates among some person-level characteristics.  

```{r process_gqs}
gq_seed <- rsp_process_gq_seed(
  pums_dat = pums_dat, 
  pumas    = pumas_use
)

gq_tgt  <- rsp_process_gq_tgt(
  gq_dat   = gq_dat[[1]], 
  fips_use = fips_use
)

gq_pop <- bind_rows(lapply(
  synth_fips,
  function(i){
    rsp_synth_gq(i, gq_seed, gq_tgt)
  }  
))

```

#### Process data to generate household and person seed and targets  
```{r process_ps_hhs}
hh_seed <- rsp_process_hh_seed(
  pums_dat         = pums_dat, 
  pumas            = pumas_use, 
  hh_vars          = hh_vars, 
  hh_income_breaks = c(-100,50,100,100000)*1000
)

hh_tgt <- rsp_process_hh_tgt(
  acs_hh_dat       = acs_dat,
  chars            = characteristics,
  fips_use         = fips_use,
  hh_income_breaks = c(-100,50,100,100000)*1000
)

p_seed <- rsp_process_p_seed(
  pums_dat     = pums_dat, 
  pumas        = pumas_use, 
  p_vars       = p_vars, 
  p_age_breaks = c(0,10,20,30,40,50,60,70,80,150)
)

p_tgt <- rsp_process_p_tgt(
  acs_p_dat       = acs_dat,
  chars            = characteristics,
  fips_use         = fips_use,
  p_age_breaks = c(0,10,20,30,40,50,60,70,80,150)
)
```

### Synthesizing population  
Now that we have the group quarters population and the household and person seed and target data, we can synthesize the population. For the most accurate population, it is recommended to synthesize at the census tract level, but for large geographies (large states or groups of states with large populations) this may lead to considerable runtime.  

```{r synth_pop, cache=TRUE}
clust <- makeCluster(detectCores()-1)
clusterExport(clust, c("synth_fips", 
                       "hh_seed", "p_seed",
                       "hh_tgt", "p_tgt",
                       "gq_seed", "gq_tgt", "gq_pop"))

clusterEvalQ(clust, library(Rsynthpops))


synth_pop <- bind_rows(parLapply(clust, 
                                 synth_fips, function(ct){
                                   rsp_synth_hhp(ct, hh_seed, p_seed, hh_tgt, p_tgt, gq_pop)
                                 }
))

stopCluster(clust)
```

```{r synth_pop_lapply, eval = FALSE, include = FALSE, echo = FALSE}
synth_pop <- lapply(synth_fips, function(ct){
  cat(ct, "\n")
  rsp_synth_hhp(ct, hh_seed, p_seed, hh_tgt, p_tgt, gq_pop)
})

```

### Validating population  

#### Global comparison of true to synthetic population  
```{r global_val, echo = FALSE}
# HHincome
hhincome_obs <- colSums(hh_tgt$HH_Income[,-1])
hhincome_gen <- synth_pop[!duplicated(synth_pop$hhid),] %>% group_by(HH_Income) %>% summarise(n = n())

hhincome_comp <- tibble(vrbl = "HH_Income",
                           lvl = as.factor(hhincome_gen$HH_Income),
                           obs = hhincome_obs,
                           gen = hhincome_gen$n)

hhincome_comp_plot <- hhincome_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = c(0.2, 0.9)) +
    scale_x_discrete(labels = c("<$50k", "$50k-$100k", ">$100k")) +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Household Type",
         y = "Count",
         fill = "")

# HHtype
hhtype_obs <- colSums(hh_tgt$HH_Type[,-1])
hhtype_gen <- synth_pop[!duplicated(synth_pop$hhid),] %>% group_by(HH_Type) %>% summarise(n = n())

hhtype_comp <- tibble(vrbl = "HH_Type",
                         lvl = as.factor(hhtype_gen$HH_Type),
                         obs = hhtype_obs,
                         gen = hhtype_gen$n)

hhtype_comp_plot <- hhtype_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_x_discrete(labels = c("Family", "non-Family")) +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Household Type",
         y = "Count",
         fill = "")


# HHsize
hhsize_obs <- colSums(hh_tgt$HH_Size[,-1])
hhsize_gen <- synth_pop[!duplicated(synth_pop$hhid),] %>% group_by(HH_Size) %>% summarise(n = n())

hhsize_comp <- tibble(vrbl = "HH_Size",
                         lvl = as.factor(hhsize_gen$HH_Size),
                         obs = hhsize_obs,
                         gen = hhsize_gen$n)

hhsize_comp_plot <- hhsize_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_x_discrete(labels = c(as.character(c(1:6)), "7+")) +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Household Size",
         y = "Count",
         fill = "")


# Grade
grade_obs <- colSums(p_tgt$Grade[,-1])
grade_gen <- synth_pop %>% group_by(Grade) %>% summarise(n = n())

grade_comp <- tibble(vrbl = "Grade",
                         lvl = as.factor(grade_gen$Grade),
                         obs = grade_obs,
                         gen = grade_gen$n)


grade_comp_plot <- grade_comp %>% 
  filter(lvl != "bb") %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Grade",
         y = "Count",
         fill = "")


# Sex
sex_obs <- colSums(p_tgt$Sex[,-1])
sex_gen <- synth_pop %>% group_by(Sex) %>% summarise(n = n())

sex_comp <- tibble(vrbl = "Sex",
                      lvl = as.factor(sex_gen$Sex),
                      obs = sex_obs,
                      gen = sex_gen$n)

sex_comp_plot <- sex_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_x_discrete(labels = c("Male", "Female")) +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Sex",
         y = "Count",
         fill = "")


# Occupational group
occ_obs <- colSums(p_tgt$Occ_Group[,-1])
occ_gen <- synth_pop %>% group_by(Occ_Group) %>% summarise(n = n())

occ_comp <- tibble(vrbl = "occ",
                      lvl = as.factor(occ_gen$Occ_Group),
                      obs = occ_obs,
                      gen = occ_gen$n)

occ_comp_plot <- occ_comp %>% 
  filter(lvl != "99") %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    annotate(geom = "text", 15,70000, label = "* Unemployed not shown") +
    labs(x = "Occupation Category",
         y = "Count",
         fill = "")


# Age Category
age_obs <- colSums(p_tgt$Age[,-1])
age_gen <- synth_pop %>% group_by(Age) %>% summarise(n = n())

age_comp <- tibble(vrbl = "age",
                      lvl = as.factor(age_gen$Age),
                      obs = age_obs,
                      gen = age_gen$n)

age_comp_plot <- age_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    scale_x_discrete(labels = c("0-9", "10-19", "20-29",
                                "30-39", "40-49", "50-59", 
                                "60-69", "70-79", "80+")) +
    labs(x = "Age Category",
         y = "Count",
         fill = "")


# Race
race_obs <- colSums(p_tgt$Race[,-1])
race_gen <- synth_pop %>% group_by(Race) %>% summarise(n = n())

race_comp <- tibble(vrbl = "Race",
                       lvl = as.factor(race_gen$Race),
                       obs = race_obs,
                       gen = race_gen$n)

race_comp_plot <- race_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    scale_x_discrete(labels = c("White", "Black", "Am. Ind\nAK nat.",
                                "Asian", "HI/PI", "Other", "2+")) +
    labs(x = "Race",
         y = "Count",
         fill = "")


# Hispanic ethnicity
eth_obs <- colSums(p_tgt$Hispanic[,-1])
eth_gen <- synth_pop %>% group_by(Hispanic) %>% summarise(n = n())

eth_comp <- tibble(vrbl = "Hispanic",
                      lvl = as.factor(eth_gen$Hispanic),
                      obs = eth_obs,
                      gen = eth_gen$n)

eth_comp_plot <- eth_comp %>% 
  pivot_longer(cols = c("gen", "obs"),
               names_to = "Obs_Gen",
               values_to = "Total") %>% 
  ggplot(aes(x = lvl, y = Total, fill = Obs_Gen)) +
    geom_col(position = "dodge") +
    theme_classic() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("#6768b2", "#64a3cd")) +
    labs(x = "Hispanic Ethnicity",
         y = "Count",
         fill = "")

# All together
gen_obs_comp <- bind_rows(
  hhincome_comp,
  hhtype_comp,
  hhsize_comp,
  grade_comp,
  sex_comp,
  occ_comp,
  age_comp,
  race_comp,
  eth_comp
) %>% 
  mutate(prop_diff = abs(obs - gen)/obs)


(hhincome_comp_plot | hhtype_comp_plot | hhsize_comp_plot) / (age_comp_plot | sex_comp_plot | grade_comp_plot) / (occ_comp_plot | race_comp_plot | eth_comp_plot)

ggsave(here::here("Plots/Overall_Validation_Add_GQ.png"),
       height = 8, width = 12, units = "in")
```

#### Deviation from attribute targets by census tract   

```{r aggregate_validation, eval=FALSE}
n_vars <- length(hh_tgt) + length(p_tgt)
n_cts  <- length(synth_cts)

val_df_tgt <- data.frame("ct"    = rep(synth_cts, times = n_vars),
                         "vrbl"  = rep(c(names(hh_tgt), names(p_tgt)), each = n_cts),
                         "pmiss" = NA)

val_df_tgt$pmiss <- apply(val_df_tgt, 1, FUN = function(i){
  ct   = as.character(i[1])
  vrbl = as.character(i[2])
  
  #cat(vrbl, "\n")
  
  if(vrbl %in% names(hh_tgt)){
    
    obs <- lapply(hh_tgt, function(i){
      i %>% filter(GEOID == ct) %>% dplyr::select(-GEOID)
    })[[vrbl]] %>% 
      pivot_longer(cols = everything(), names_to = vrbl, values_to = "n") %>% 
      group_by(!!as.name(vrbl)) %>% 
      summarise(n = sum(n))
    
    # One obs per household
    ct <- synth_pop %>% filter(GEOID == ct)
    ct_hhs <- ct[!duplicated(ct$hhid),]

    gen <- ct_hhs %>% 
      group_by(!!as.name(vrbl)) %>% 
      dplyr::count()
    
    class(gen[[vrbl]]) <- class(obs[[vrbl]])
    
    jnd <- obs %>% 
      left_join(gen, by = vrbl, suffix = c("_obs", "_gen")) %>% 
      mutate(diff = abs(n_obs - n_gen))
    
    out <- sum(jnd$diff, na.rm = T)/sum(jnd$n_obs)
    
  } else if(vrbl %in% names(p_tgt)){
    
    obs <- lapply(p_tgt, function(j){
      j %>% filter(GEOID == ct) %>% dplyr::select(-GEOID)
    })[[vrbl]] %>% 
       pivot_longer(cols = everything(), names_to = vrbl, values_to = "n")
    
    gen <- synth_pop %>% 
      filter(GEOID == ct) %>% 
      group_by(!!as.name(vrbl)) %>% 
      dplyr::count()
    
    class(gen[[vrbl]]) <- class(obs[[vrbl]])

    jnd <- obs %>% 
      left_join(gen, by = vrbl, suffix = c("_obs", "_gen")) %>% 
      mutate(diff = abs(n_obs - n_gen))

    out <- sum(jnd$diff, na.rm = T)/sum(jnd$n_obs)

  } else {
    
    stop("Variable has no match in target data")
    
  }
  
  return(out)
  
})

val_df_tgt %>% 
  ggplot(aes(x = vrbl, y = pmiss, fill = vrbl)) +
    geom_violin() +
    scale_y_continuous(breaks = seq(0,2, by = 0.1)) +
    scale_x_discrete(labels = c("HH_Income" = "HH\nIncome",
                                "HH_Type"   = "HH\nType",
                                "HH_Size"   = "HH\nSize",
                                "Grade"    = "Grade",
                                "Sex"      = "Sex",
                                "Occ_Group"= "Occp\nGroup",
                                "Age"  = "Age\nCat",
                                "Race"     = "Race",
                                "Hispanic" = "Hispanic")) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(x = "Variable",
         y = "Normalized L1",
         title = "L1 normalized error distribution across CTs")

ggsave(here::here("Plots/PctError_Distn_Validation.png"),
       height = 4, width = 5, units = "in")

high_err_cts <- val_df_tgt %>% 
  filter(pmiss > 0.5) %>% 
  pull(ct) %>% 
  unique()

val_df_tgt %>% 
  filter(!ct %in% high_err_cts) %>% 
  ggplot(aes(x = vrbl, y = pmiss, fill = vrbl)) +
    geom_violin() +
    scale_y_continuous(breaks = seq(0,2, by = 0.1)) +
    scale_x_discrete(labels = c("HH_Income" = "HH\nIncome",
                                "HH_Type"   = "HH\nType",
                                "HH_Size"   = "HH\nSize",
                                "Grade"    = "Grade",
                                "Sex"      = "Sex",
                                "Occ_Group"= "Occp\nGroup",
                                "Age"  = "Age\nCat",
                                "Race"     = "Race",
                                "Hispanic" = "Hispanic")) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(x = "Variable",
         y = "Normalized L1",
         title = "L1 normalized error distribution across CTs",
         subtitle = " *Large error CTs omitted")

ggsave(here::here("Plots/PctError_Distn_Validation_rmv_high_error.png"),
       height = 4, width = 5, units = "in")

```
